Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-09-03 manuscript.qmd
make model vars dynamic

The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated. For example, midline extension was present in {{< var data.advanced_with_midext_percent >}}% of advanced T-category tumors but only in {{< var data.early_with_midext_percent >}}% of early T-category tumors. The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors. However, the bottom right panel of @fig-data-strat indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors. The figure considers only patients with lateralized tumors not extending over the midline. For early T-category and healthy ipsilateral levels I-V, only {{< var data.early_ipsin0_cII_percent >}}% ({{< var data.early_ipsin0_cII_match >}} out of {{< var data.early_ipsin0_cII_total >}} patients) shows involvement of contralateral level II. This increases to {{< var data.early_ipsiII_cII_percent >}}% ({{< var data.early_ipsiII_cII_match >}} out of {{< var data.early_ipsiII_cII_total >}} patients) if ipsilateral level II is involved, to {{< var data.early_ipsiIIandIII_cII_percent >}}% ({{< var data.early_ipsiIIandIII_cII_match >}} out of {{< var data.early_ipsiIIandIII_cII_total >}} patients) if ipsilateral levels II and III are involved, and to {{< var data.late_ipsiIIandIII_cII_percent >}}% ({{< var data.late_ipsiIIandIII_cII_match >}} out of {{< var data.late_ipsiIIandIII_cII_total >}} patients) for advanced T-category with ipsilateral levels II and III involved.

@@

For example, we may be interested in the prevalence of contralateral LNL II involvement (i.e., contra LNL II "involved" and all other LNLs "masked") under the scenario of early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$). To compute this prevalence in the data, we select all patients of this scenario (in our data, this amounts to 379<!-- TODO: dyamically compute this --> patients). Of those, 28<!-- TODO: dynamically compute this --> were found to harbor metastases in their contralateral LNL II. Therefore, the prevalence is 7%<!-- TODO: dynamically compute this -->.

@@

In this subsection we illustrate parts of the mathematical framework defined in the previous section. In @fig-model-midext-evo (top panel), we visualize the prior distribution over diagnose times $P(t)$. By parameter choice, early T-category tumors are diagnosed on average after 3 time steps. Advanced T-category tumors are diagnosed on average after 4.5 time steps ($p_\text{adv.}=0.45$) <!-- TODO: make dynamic -->, a parameter learned from the data to describe the higher involvement for advanced T-category tumors. The probability per time step for the tumor to grow over the midline, $p_\epsilon = 0.0816$ <!-- TODO: make dynamic --> yields the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line). The parameter is adjusted such that the model yields the correct proportion of patients with tumors extending over the midline seen in the dataset. The bottom panel in @fig-model-midext-evo shows the joint probability $P(\epsilon, t)$ to be diagnosed at time step $t$ with a given state of midline extension and T-category.

@@

@Fig-model-risks (left panel) shows the predicted risk of occult disease in contralateral LNL II. The most important variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization. A patient with a clearly lateralized early T-category tumor and a clinically N0 neck is predicted to have a 1-2% <!-- TODO: make this dynamic, and include this scenario in the plot --> risk for occult disease in contralateral LNL II. For a patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to almost 7%. <!-- TODO: make this dynamic -->

>>>>

The three risk factors for contralateral involvement, midline extension, T-category, and ipsilateral involvement are correlated. For example, midline extension was present in {{< var data.late_with_midext_percent >}}% of advanced T-category tumors but only in {{< var data.early_with_midext_percent >}}% of early T-category tumors. The higher prevalence of contralateral metastases in advanced T-category tumors may thus be partially explained by a higher rate of midline extension in these tumors. However, the bottom right panel of @fig-data-strat indicates that advanced T-category and severe ipsilateral involvement do represent additional risk factors. The figure considers only patients with lateralized tumors not extending over the midline. For early T-category and healthy ipsilateral levels I-V, only {{< var data.early_ipsin0_cII_percent >}}% ({{< var data.early_ipsin0_cII_match >}} out of {{< var data.early_ipsin0_cII_total >}} patients) shows involvement of contralateral level II. This increases to {{< var data.early_ipsiII_cII_percent >}}% ({{< var data.early_ipsiII_cII_match >}} out of {{< var data.early_ipsiII_cII_total >}} patients) if ipsilateral level II is involved, to {{< var data.early_ipsiIIandIII_cII_percent >}}% ({{< var data.early_ipsiIIandIII_cII_match >}} out of {{< var data.early_ipsiIIandIII_cII_total >}} patients) if ipsilateral levels II and III are involved, and to {{< var data.late_ipsiIIandIII_cII_percent >}}% ({{< var data.late_ipsiIIandIII_cII_match >}} out of {{< var data.late_ipsiIIandIII_cII_total >}} patients) for advanced T-category with ipsilateral levels II and III involved.

@@

For example, we may be interested in the prevalence of contralateral LNL II involvement (i.e., contra LNL II "involved" and all other LNLs "masked") under the scenario of early T-category (T0-T2) and no midline extension ($\epsilon=\texttt{False}$). To compute this prevalence in the data, we select all patients of this scenario (in our data, this amounts to {{< var data.early_nomidext_cII_total >}} patients). Of those, {{< var data.early_nomidext_cII_match >}} were found to harbor metastases in their contralateral LNL II. Therefore, the prevalence is {{< var data.early_nomidext_cII_percent >}}%.

@@

In this subsection we illustrate parts of the mathematical framework defined in the previous section. In @fig-model-midext-evo (top panel), we visualize the prior distribution over diagnose times $P(t)$. By parameter choice, early T-category tumors are diagnosed on average after {{< var model.early_expected_time >}} time steps. Advanced T-category tumors are diagnosed on average after {{< var model.late_expected_time >}} time steps, because the parameter $p_\text{adv.}$ was learned to be {{< var model.params_percent.late_p >}}% for advanced T-category tumors. The probability per time step for the tumor to grow over the midline ($p_\epsilon$) was on average {{< var model.params_percent.midext_prob >}}%. Using this, one can compute the probability of midline extension $P(\epsilon \mid t)$ at a given time step $t$ (red line in @fig-model-midext-evo). The bottom panel in this figure shows the joint probability $P(\epsilon, t)$ to be diagnosed at time step $t$ with a given state of midline extension and T-category.

@@

@Fig-model-risks (left panel) shows the predicted risk of occult disease in contralateral LNL II. The most important variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization. A patient with a clearly lateralized early T-category tumor and a clinically N0 neck is predicted to have a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II. For a patient with a clearly lateralized early T-category tumor but *with* mid-sagittal extension, the risk increases to almost 7%. <!-- TODO: make this dynamic -->
